### 2011-10-17 trunk/papers/smectics/smectics.tex
Conclusions and references edited
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1464 40876441-3496-0410-a613-a2fca35b2ae2

Instances of soft complex fluids abound. Examples include colloidal
suspensions, where particles or droplets are suspended in a viscous
fluid, living tissues, and liquid crystalline fluids. Well known
examples of colloidal suspensions include paint, cosmetics, and
food products such as ketchup, mayonnaise, and yogurt.
The intracellular network of actin fibres constituting the
cytoskeleton of mammalian cells is a polymeric fluid.
The same is true of the DNA making up the bacterial nucleoid and
eukaryotic chromosomes, and the viral genome of bacteriophages,
where enhanced confinement imparts a strong liquid crystalline
character to the system.
Liquid crystalline fluids themselves, with or without colloidal
inclusions, are of of wide technological importance in devices
such as electronic displays and sensors.

@@

In order to ensure thermodynamic stability the parameters $b$ and $c$ have to be positive.
For $a>0$ there is only one unique solution $\phi\equiv0$ which minimises the functional Eq. \ref{FE-functional} and wich reprsents the the disordered state.
For $a<0$ a microphase separation occurs.
Depending on the sign of $\kappa$ either two homogenous phases coexist ($\kappa>0$) or a transition to an ordered lamellar phase state takes place ($\kappa<0$) as the formation of interfaces become energetically favourable.
The lamellar width is approximately given by $2\pi /\sqrt{|\kappa|/2 c}$, whereas the interfacial width is roughly $\sqrt{2|\kappa|/|a|}$. 
This entails that by varying $a$ (and $b\sim |a|$) by about on decade, the segregation of the two phases varies by about a factor $3$.\\
The dynamics of the binary fluid is governed by an advection-diffusion equation with a diffusive current that is proportional to the chemical potential $\mu$, the functional derivative of ${\cal F}$ with respect to $\phi$.

@@

with $\phi({\mathbf k})$ being the normalised Fourier-transform of the order parameter.
The transform itself was calculated using the FFTW-library \cite{FFTW}.\\
For the analysis of the relation between rheological quantities and the morphology of the system it was necessary to find a means to quantify the number of defects.
For this purpose we defined an alignment factor $X_i$, which measures the average projection of the normalised order parameter gradients onto the gradient at site $i$.
More explicitly the alignment factor $X_i$ is given by

\begin{equation}\label{alignment}
X_i=\frac{1}{N_{nn}} \sum_{j=1,N_{nn}}\left(\frac{\nabla \phi_i \cdot \nabla \phi_j}{|\nabla\phi_i||\nabla\phi_j|}\right)^2. 
\end{equation}

$N_{nn}$ is the number of nearest neighbours ($N_{nn}=8$ in 2D and $N_{nn}=26$ in 3D).
For smooth alignment and small curvature of the lamellae $X_i$ is very close to unity.
At the endpoints of the lamellae or in regions where their orientation changes abruptly it departs form unity and takes smaller positive values.
By setting a threshold $\xi$ it is possible to detect domains where the order parameter structure deviates preceivably from what would be recognised as an aligned configuration exhibiting lamellar order.
We found a threshold value of $\xi=0.8$ to be well suitable for our requirements.
Unfortunately, this defect detecting protocol is not able to assign only single sites to defect regions as naturally more than one site would have an alignment factor below the threshold.
However, as the number of sites per defect is on average more or less constant throughout the sample, this protocol provides still a useful way to quantify the degree of order and alignment in various flow conditions.\\
By normalising the number of defects sites with $X_i$ below the threshold a defect density can be defined:

\begin{equation}\label{defect density}
\rho_D=\frac{1}{N} \sum_{i=1,N} \theta(\xi-X_i).  
\end{equation}

The function $\theta$ is the Heaviside function and $N=N_x\times N_y \times N_z$ the total number of sites in the simulation.
The exact value of the defect density clearly depends on the lamellar width and of course on the threshold $\xi$ for the alignment parameter.

@@

Our discussion will be in two parts, beginning with simulations in two dimensions for systems A-2D and B-2D.
The system size was $N_x\times N_y\times N_z=512\time1024\times 1$ and $N_x \times N_y \times N_z=256\times512\times 1$ respectively for the longtime runs in the nonlinear rheological regime.
Phase segregation was controlled by modifying the Landau parameters $a$ and $b$ in the bulk free energy term of Eq. \ref{FE-functional} while keeping the parameters $\kappa$ and $c$ controlling the gradient terms constant.
By this we obtained a slightly different stress response with the system with lower segregation level appearing somwhat sofer.
The lamellar distance was identical in both cases.\\
The second part of our discussion consists of simulations in three dimensions, referred to as A-3D and C-3D using identical parameters to system A-2D and C-2D and $N_x\times N_y\times N_z=128\times256\times128$.
Note that velocity gradient and flow direction are oriented along the x- and y-axis, respectively, which is the {\it vertical} and the {\it horizontal} dimension in the 2D-plots. 
Table \ref{tab1} gives an overview of all simulation parameters.

@@

Lamellar phases occur within the Landau-Brazovskii free energy functional approach Eq. \ref{FE-functional} due to the interplay of gain in free energy by virtue of phase separation and loss due to local curvature.
Hence, a range of parametrizations can produce lamellar phases with almost identical outcome.
In previous studies \cite{Kendon01,Xu03, Xu06b} parameters have been identified which cover the relevant morphology of lamellar phases and to produce stable outcomes. 
We performed simulations for Landau parameters $5\cdot10^{-4}\le-a=b\le5\cdot10^{-5}$ and viscosities $8.33\cdot10^{-4}\le\eta\le 0.833$.
Reasonable numerical stability was attained for all Landau parameters in this range and for viscosities $\eta\ge 8.33\e{-3}$.
The morphology of the lamellar structure and the kinetics of the spinodal decomposition depended on the specific choice though.
Smaller Landau parameters slowed down the ripening of the lamellar structure, which is consistent with the fact that systems with lower segregation have lower average order parameters and are generally closer to the critical point.
Higher viscosities lead to shorter interfaces, whereas lower values typically invoke longer lamellae.
This effect has been reported before \cite{Gonnella97}.
It turned out to be more pronounced for low segregation, although not as quite to the same extent as in the above reference.
In shear flow high values of $\eta$ entailed more inhomogenous lamellae in terms of shape and thickness.\\
In the following we will show characteristic results for viscosity $\eta=8.33\e{-2}$ and the systems A-2D and B-2D, respectively.
The top picture in Fig. \ref{fig1} shows the order parameter $\phi$ at the end of the equilibration phase prior to shearing.
The equilibrium structure is completely isotropic and exhibits a lamellar spacing $L$ of roughly 10 lattice sites.
If shear flow is imposed, the lamellae start to align in flow direction, which induces a morphological changes to the structure.

@@

@@

The flow direction and the direction of the velocity gradient are {\it vertical} x- and {\it horizontal} y-direction.
The relative flow imposed by the Lees-Edwards planes is always in positive y-direction in the upper half and in negative y-direction in the lower half of the images.
After a short transient phase during which the order parameter remains rather homogenous, regions of reduced and increased lamellar width emerge and the spacing begins to deviate from its well-defined equilibrium value.
These regions are advected with the flow while they continuously undergo affine transformations.
The centre picture in Fig. \ref{fig1} shows an example at an intermediate time step at shear rate $\gd= 1\e{-4}$.
In a previous work on binary fluids and lamellar systems in shear flow \cite{Xu06b} another shear-induced structure was found.
However, the reported results differ from our observations in the following aspects.
Xu et al. shear-induced structures emerge at the centre and remain localised, while ours do occur in band-like patterns that are wrapped around the periodic boundaries of our system.
Xu et al. also reported a shear banding region with reduced local shear rate, which is absent in our case.
These differences may be simply a direct consequence of the different boundary conditions that were applied.
While we applied Lees-Edwards BCs to all quantities, Xu and co-workers seem to have used a kind of hybrid scheme.

\begin{figure}[htp!]
\centering
\hspace*{-0.1cm}\includegraphics[angle=0,width=0.5\textwidth]{phi_run704_400.jpg}
\includegraphics[angle=0,width=0.49\textwidth]{shear_str_run704_400.jpg}
\includegraphics[angle=0,width=0.49\textwidth]{N1_run704_400.jpg}
\caption{Order parameter $\phi({\mathbf r})$ (top), shear stress $\sigma_{xy}({\mathbf r})$ (centre) and first normal stress differences $N_1({\mathbf r})$ (bottom) of system A-2D at $t=4\e{5}$}
\label{fig3}
\end{figure}

\begin{figure}[htp!]
\centering
\includegraphics[angle=0,width=0.33\textwidth]{ck_run703_320.jpg}
\includegraphics[angle=0,width=0.33\textwidth]{ck_run704_500.jpg}
\includegraphics[angle=0,width=0.33\textwidth]{ck_run705_1280.jpg}
\caption{Structure factor $C({\bm k})$ of system A-2D: Shown are states after equilibration at $t=3.2\e{5}$ (left), at $\dot{\gamma}=1\cdot10^{-4}$ and $t=5\e{5}$ (centre) and after switch-off at $t=1.28\e{6}$ (right). The sections shown correspond to wave vectors on the interval $k_x, k_y \in [-\pi/2 l,\pi/2 l]$.}
\label{fig4}
\end{figure}

To our understanding Lees-Edwards BC were applied to the order parameter only, whereas solid moving walls were placed in the LB-part at the upper and lower boundaries.
It might be that in \cite{Xu06b} actually not bulk flow behaviour was simulated, but rather flow in a confined geometry. 
This could also explain why Xu et al. always observe a shear-induced structure at the centre of the system.\\ 
After the shear flow has been switched off the lamellar spacing became again more homogenous and the equilibrium value was restored with expanded regions contracting and vice versa.
The final state after the relaxation has virtually come to a standstill is shown in the bottom picture of Fig. \ref{fig1}. 
Most of the lamellae remain oriented along the flow direction, but depending on the local expansion and compression level during flow some regions may have large angles after relaxation. 
The length of the interfaces varies considerably, which is another residue of the inhomogeneities present in shear flow.\\
Fig. \ref{fig2} shows equivalent states for system B-2D with lower segregation.
All generic results of Fig. \ref{fig1} are the same, but the interfaces appear to be slightly longer.
Consequently, the lamellar alignment in flow direction after switch off is also more pronounced.\\ 
In Fig. \ref{fig3} we suggest a relation between the inhomogeneous morphology in the flowing state and typical rheological quantities.
The pictures show the order parameter $\phi({\bm r})$, shear stress density $\sigma_{xy}({\bm r})$ and first normal stress difference density $N1({\bm r})=\sigma_{yy}({\bm r})-\sigma_{xx}({\bm r})$ of system A-2D in steady shear flow at shear rate $\dot{\gamma}=1\e{-4}$.
In regions where the lamellar width is larger than in the equilibrium state shear stresses and first normal stress differences show low positive or negative values.
Large positive values of both quantities coincide with regions where the lamellae are under compression.
Peak values in LBU are in the region of $-2\e{-4}$ and $3\e{-4}$ for shear stress and  $-3.5\e{-4}$ and $7\e{-4}$ for the normal stress differences.
Despite being locall negative, the ensemble averages of these quantities are always positive.
Very similar results with smaller peak magnitudes are obtained for system B-2D.
Fig. \ref{fig4} shows structure factors $C({\bm k})$ for the same states as Fig. \ref{fig1}.
Initially isotropic after spinodal decomposition and equilibration, $C({\bm k})$ has a distinct peak at $|{\bm k}|\simeq \pm0.2\pi/l$.
In steady shear flow the structure factor develops two distinct but rather broad peaks at $k_x\simeq\mp 0.012\pi/l, k_y\simeq\pm 0.22 \pi/l$.
It is interesting to see that the inhomogenous morphology of the lamellar structure is also clearly encoded in the structure factor. 
The Fourier-vector at the peak has a small component in x-direction, indicating the average tilt of the lamellae away from the flow direction in the direction of extensional flow.
The position of the peak is almost constant throughout the entire simulation and the average lamellar spacing decreases from about 10 lattice sites in equilibrium to about 8.5 lattice sites.
If the shear flow is switched off the system adopts again its equilibrium value and the small tilt of the signal vanishes completely, indicating that this is a purely dynamical phenomenon. 
The intensity of the peak is slightly larger due to the fact that the interfaces remain oriented along the flow direction.
By introducing the alignment factor $X_i$ in Eq. \ref{alignment} we suggest a method to identify and quantify defects in the lamellar structure.
Fig. \ref{fig5} shows an example of the time evolution of system A-2D at low shear rate.
All lattice sites with $X_i\le0.8$ are marked in red and they obviously coincide with lamellar endpoints or local disturbances.
The top picture gives the situation at the lowest applied shear rate $\dot{\gamma}=1.25\e{-5}$, whereas the system has been pre-sheared at consecutively decreasing shear rates between $\gd=1\e{-4}$ and $2.5\e{-5}$.
There are regions which show some sort of flow alignment and others which don't.
The bottom picture of Fig. \ref{fig5} shows a later state with a total strain that has increased by another $10^{5}\%$. 
Obviously many defects have annealed and the interfaces are well aligned with the flow direction.

\begin{figure}[htp!]
\centering
\includegraphics[angle=0,width=0.5\textwidth]{phi_defects_run774_5120.jpg}
\includegraphics[angle=0,width=0.5\textwidth]{phi_defects_run774_10240.jpg}
\caption{Defect structure of system A-2D: The pictures show results at timestep $t=5.12\e{6}$ (top) and $1.024\e{7}$ (bottom) with an imposed shear rate $\dot{\gamma}=1.25\cdot10^{-5}$. The system has been pre-sheared at various higher, subsequently lowered shear rates ranging from $\dot{\gamma}=2\e{-4}$ to the above value (see Tab.\ref{tab1}). Sites where the alignment factor $X_i$ was smaller than $0.8$ are highlighted red.}
\label{fig5}
\end{figure}

Defect annihilation to this large extent is only found below a specific shear rate.
A prerequisite seems to be that the defects can smootly approach each other and cancel, which is not possible above the instability as the lamellae buckle similar to \cite{Gonnella98}, show a tendency to trip over themselves which results in the creation of new defects.
The critical value depends on the segregation and differs in systems A-2D and B-2D.
It is roughly between $\dot{\gamma}=2.5-5\e{-5}$ in A-2D and around $\dot{\gamma}=1-2\e{-4}$ in B-2D.\\
Fig. \ref{fig6} shows the  complete time evolution of the defects density $\rho_D$ of system A-2D for a sequence of decreasing shear rates starting at $\gd=2\e{-4}$ and ranging down to $\gd=2.5\e{-5}$.
The runs with the two highest shear rates were started from the equilibrium configuration, whereas all others were pre-sheared using the final configuration of the preceding run as initial configuration in order to save computational rexources.
The total strain between the beginning and end of every individual run was at least $10^{5}\%$.
Interestingly, while stepping down in shear rate both systems A-2D and B-2D anneal defects almost instantaneously.
If the shear rate is turned up again (not shown), the lamellae buckle and new defects recur immediately.

\begin{figure}[htp!]
\centering
\includegraphics[angle=0,width=0.5\textwidth]{defect_density_5e-4.pdf}
\caption{Defect density $\rho_D$ vs time: The main picture shows data for system A-2D of a sequence of simulations with consecutively lowered shear rate $\dot{\gamma}=2\e{-4}$ (red solid), $1\e{-4}$ (green long-dashed),  $5\e{-5}$ (blue short-dashed),  $2.5\e{-5}$ (magenta dotted) and  $1.25\e{-5}$ (cyan dashed-dotted). Horizontal lines indicate time averages of the particular quantities, where we assume the system has reached a steady state. The inset gives data for the system A-3D and shear rates $\dot{\gamma}=2\cdot10^{-4}, 1\cdot10^{-4}$ and $5\cdot10^{-5}$.} 
\label{fig6}
\end{figure}

@@

This feature which appears somewhat erratic at first has actually a simple explanation. 
The time sequence of the defect density in Fig. \ref{fig6} reveals that the rise in $N_1$ is correlated with a drop in $\rho_D$ and linked to the defect annihilation.
When two defects meet and cancel the lamellar width increases locally by a certain amount.
This causes frustration in the system as the incommensurability between the layers is not large enough to create a new lamellar layer and allow the system to snap back towards its equilibrium width.
Hence the normal stresses $\sigma_{xx}$ become increasingly smaller and may even attain negative values, leading to gradually larger values of $N_1$.
This explanation is further sustained by the fact that only $\sigma_{xx}$ seems to evolve anomalously slowly and all other components of the stress tensor reach their steady state values more quickly.
We expect that these anomalous trends become considerably smaller in larger systems.
System B-2D behaves differently in the sense that $N_1$ does not follow this trend but rather continues to fall for lower shear rates. 
This is likely to be a direct consequence of the lower segregation in system B-2D, which gives rise to positive normal stresses $\sigma_{xx}$ preventing $N_1$ from increasing.\\
In Fig. \ref{fig11} the shear rate dependence of $\sigma_{xy}$ and $N_1$ is shown. 
System A-2D gives rise to shear thinning with the average shear stress of systems A-2D following a power law $\sigma_{xy}=\dot{\gamma}^m$ with exponent $m\simeq 0.87$.
For system B-2D the best fit is actually obtained for an exponent $m=1$, indicating more or less constant viscosity and no pronounced shear thinning.
This is because due to the decreasing number of defects at shear rates below $\dot{\gamma}\le2\e{-4}$ the chemical stress $P^{chem}_{xy}$ almost vanishes and the total stress is dominated by the hydrodynamic part.\\

>>>>

Instances of soft complex fluids are abound in our everyday world. 
Examples include colloidal suspensions, 
where particles or droplets are suspended in a viscous
fluid, living tissues, and liquid crystalline fluids. Well known
examples of colloidal suspensions include paint, cosmetics, and
food products such as ketchup, mayonnaise, and yogurt.
The intracellular network of actin fibres constituting the
cytoskeleton of mammalian cells is a polymeric fluid.
The same is true of the DNA making up the bacterial nucleoid and
eukaryotic chromosomes, and the viral genome of bacteriophages,
where enhanced confinement imparts a strong liquid crystalline
character to the system.
Liquid crystalline fluids themselves, with or without colloidal
inclusions, are of of wide technological importance in devices
such as electronic displays and sensors.

@@

In order to ensure thermodynamic stability the parameters $b$ and $c$ have to be positive.
For $a>0$ there is only one unique solution $\phi\equiv0$ which minimises the functional Eq. \ref{FE-functional} and wich reprsents the the disordered state.
For $a<0$ a microphase separation occurs.
Depending on the sign of $\kappa$ either two homogenous phases coexist ($\kappa>0$) or a transition to an ordered lamellar phase state takes place ($\kappa<0$) as the formation of interfaces become energetically favourable.
The lamellar width is approximately given by $2\pi /\sqrt{|\kappa|/2 c}$. 
By varying the Landau-parameters $a$ (and $b\sim |a|$) we change the distance from the critical point and the magnitude of the order-parameter.
The segregation between both phases was found to dependent neither solely on $a,b$ nor only on $\kappa,c$. 
It is rather the magnitude of all parameters, which has to be similar in order to produce a lamellar phase, that controls the level of segregation.\\
The dynamics of the binary fluid is governed by an advection-diffusion equation with a diffusive current that is proportional to the chemical potential $\mu$, the functional derivative of ${\cal F}$ with respect to $\phi$.

@@

with $\phi({\mathbf k})$ being the normalised Fourier-transform of the order parameter.
The transform itself was calculated using the FFTW-library \cite{FFTW}.\\
For the analysis of the relation between rheological quantities and the morphology of the system it was necessary to find a means to quantify the number of defects.
For this purpose we defined an alignment function $X({\mathbf r})$, which measures the average projection of the normalised order parameter gradients onto the gradient at a specific site.
More explicitly the alignment function $X$ is given by

\begin{equation}\label{alignment}
X({\mathbf r})=\frac{1}{V'({\mathbf r})} \int_{V'({\mathbf r})} d{\mathbf r'} \left(\frac{\nabla \phi ({\mathbf r}) \cdot \nabla \phi ({\mathbf r'})}{|\nabla\phi ({\mathbf r}) ||\nabla\phi ({\mathbf r'}) |}\right)^2. 
\end{equation}

$V'$ is a small volume that contains $\mathbf r$, which in practice can be chosen to contain  and its nearest neighbours.
For smooth alignment and small curvature of the lamellae $X$ is very close to unity.
At the endpoints of the lamellae or in regions where their orientation changes abruptly it departs form unity and takes smaller positive values.
By setting a threshold $\xi$ it is possible to detect domains where the order parameter structure deviates preceivably from what would be recognised as an aligned configuration exhibiting lamellar order.
We found a threshold value of $\xi=0.8$ to be well suitable for our requirements.
Unfortunately, this defect detecting protocol is not able to assign only single sites to defect regions as naturally more than one site would have an alignment function below the threshold.
However, as the number of sites per defect is on average more or less constant throughout the sample, this protocol provides still a useful way to quantify the degree of order and alignment in various flow conditions.\\
By normalising the number of defects sites with $X$ below the threshold a defect density can be defined:

\begin{equation}\label{defect density}
\rho_D({\mathbf r})=\frac{1}{V} \int d{\mathbf r'} \theta(\xi-X(\mathbf r')).  
\end{equation}

The function $\theta$ is the Heaviside function and $V$ is the volume of the entire system.
The exact value of the defect density clearly depends on the lamellar width and of course on the threshold $\xi$ for the alignment parameter.

@@

Our discussion will be in two parts, beginning with simulations in two dimensions for systems A-2D and B-2D.
The system size was $N_x\times N_y\times N_z=512\times 1024\times 1$ and $N_x \times N_y \times N_z=256\times512\times 1$ respectively for the longtime runs in the nonlinear rheological regime.
The distance from the critical point was controlled by modifying the Landau parameters $a$ and $b$ in the bulk free energy term of Eq. \ref{FE-functional} while keeping the parameters $\kappa$ and $c$ controlling the gradient terms constant.
By this we obtained a slightly different stress response with the system with lower order appearing somwhat sofer.
The lamellar distance was identical in both cases.\\
The second part of our discussion consists of simulations in three dimensions, referred to as A-3D and C-3D using identical parameters to system A-2D and C-2D and $N_x\times N_y\times N_z=128\times256\times128$.
Note that velocity gradient and flow direction are oriented along the x- and y-axis, respectively, which is the {\it vertical} and the {\it horizontal} dimension in the 2D-plots. 
Table \ref{tab1} gives an overview of all simulation parameters.

@@

Lamellar phases occur within the Landau-Brazovskii free energy functional approach Eq. \ref{FE-functional} due to the interplay of gain in free energy by virtue of phase separation and loss due to local curvature.
Hence, a range of parametrizations can produce lamellar phases with almost identical outcome.
In previous studies \cite{Kendon01,Kumaran2001,Xu03, Xu06b,Kumaran2011} parameters have been identified which cover the relevant morphology of lamellar phases and to produce stable outcomes. 
We performed simulations for Landau parameters $5\cdot10^{-4}\le-a=b\le5\cdot10^{-5}$ and viscosities $8.33\cdot10^{-4}\le\eta\le 0.833$.
Reasonable numerical stability was attained for all Landau parameters in this range and for viscosities $\eta\ge 8.33\e{-3}$.
The morphology of the lamellar structure and the kinetics of the spinodal decomposition depended on the specific choice though.
Smaller Landau parameters slowed down the ripening of the lamellar structure, which is consistent with the fact that systems B has a lower average order parameter and is closer to the critical point.
Higher viscosities lead to shorter interfaces, whereas lower values typically yield longer lamellae.
This effect has been reported before \cite{Gonnella97} and turned out to be more pronounced for the softer system B-2D, although not as quite to the same extent as in the former case.
In shear flow high values of $\eta$ entailed more inhomogenous lamellae in terms of shape and thickness.\\
In the following we will show characteristic results for viscosity $\eta=8.33\e{-2}$ and the systems A-2D and B-2D, respectively.
The top picture in Fig. \ref{fig1} shows the order parameter $\phi$ at the end of the equilibration phase prior to shearing.
The equilibrium structure is completely isotropic and exhibits a lamellar spacing $L$ of roughly 10 lattice sites.
If shear flow is imposed, the lamellae start to align in flow direction, which induces a morphological changes to the structure.

@@

The flow direction and the direction of the velocity gradient are {\it vertical} x- and {\it horizontal} y-direction.
The relative flow imposed by the Lees-Edwards planes is always in positive y-direction in the upper half and in negative y-direction in the lower half of the images.
After a short transient phase during which the order parameter remains rather homogenous, expanded and compressed regions emerge with a lamellar spacing that deviates from the equilibrium value.
This is depicted in the centre picture of Fig. \ref{fig1} for shear rate $\gd= 1\e{-4}$.
The inhomogeneities remain present in the steady state, are advected by the flow and undergo continuously affine transformations.
A buckling instability above a critical shear rate has been described earlier by Gonnella et al. \cite{Gonnella98} and Kumaran et al. \cite{Kumaran2001}.
It was suggested that the density of defects defects get introduced into a lamellar liquid crystal due to dilation and annealing occurs due to application of shear.
This would lead to irregularities in the sample a dynamical process where the steady state defect density is determined by balance between the rate of introduction and rate of annealing of defects.
In a more recent paper Kumaran et al. state that they observe similar inhomogeneities only above a certain system size \cite{Kumaran2011}, which agrees well with our observation.
We found that shear alignment is more likely to take place in smaller systems of $N_x\times N_y\times N_z=128\times 256\times 1$.
However, this trend is not necessarily followed at higher shear rates, a fact that we will discuss later in connection with our results in three dimensions.\\
Kumaran et al. also identified a second mechanism of defect creation that is due to layer compression and splitting.
Although we have not checked this in detail, this is also very likely the case in our results as the lamellar pattern is strongly compressed and heterogenous as shown in Fig. \ref{fig1}.\\
After the shear flow has been switched off the lamellar spacing became again more homogenous and the equilibrium value was restored with expanded regions contracting and vice versa.
The final state after the relaxation has virtually come to a standstill is shown in the bottom picture of Fig. \ref{fig1}. 
Most of the lamellae remain oriented along the flow direction, but depending on the local expansion and compression level during flow some regions may have large angles after relaxation. 
The length of the interfaces varies considerably, which is another residue of the inhomogeneities present in shear flow.\\
Fig. \ref{fig2} shows equivalent states for system B-2D, which is closer to the critical point.
All generic results of Fig. \ref{fig1} are the same, but the interfaces appear to be slightly longer.
Consequently, the lamellar alignment in flow direction after switch off is also more pronounced.\\ 
Another shear-induced structure in binary lamellar systems was reported in \cite{Xu06b}.
The reported results differ from our observations in some aspects.
Xu et al. shear-induced structures emerge at the centre and remain localised, while ours do occur in band-like patterns that are wrapped around the periodic boundaries of our system.
Xu et al. also reported a shear banding region with reduced local shear rate, which is absent in our case.
These differences may be simply a direct consequence of the different boundary conditions that were applied.
While we applied Lees-Edwards BCs to all quantities, Xu and co-workers seem to have used a kind of hybrid scheme.\\
To our understanding Lees-Edwards BC were applied to the order parameter only, whereas solid moving walls were placed in the LB-part at the upper and lower boundaries.
It might be that in \cite{Xu06b} actually not bulk flow behaviour was simulated, but rather flow in a confined geometry. 
This could also explain why Xu et al. always observe a shear-induced structure at the centre of the system.\\

@@

By introducing the alignment function $X$ in Eq. \ref{alignment} we suggest a method to identify and quantify defects in the lamellar structure.
Fig. \ref{fig3} shows an example of the time evolution of system A-2D at low shear rate.
All lattice sites with $X({\mathbf r})\le0.8$ are marked in red and they obviously coincide with lamellar endpoints or local disturbances.
The top picture gives the situation at the lowest applied shear rate $\dot{\gamma}=1.25\e{-5}$, whereas the system has been pre-sheared at consecutively decreasing shear rates between $\gd=1\e{-4}$ and $2.5\e{-5}$.
There are regions which show some sort of flow alignment and others which don't.
The bottom picture of Fig. \ref{fig3} shows a later state with a total strain that has increased by another $10^{5}\%$. 
Obviously many defects have annealed and the interfaces are well aligned with the flow direction.

\begin{figure}[htp]
\centering
\includegraphics[angle=0,width=0.5\textwidth]{phi_defects_run774_5120.jpg}
\includegraphics[angle=0,width=0.5\textwidth]{phi_defects_run774_10240.jpg}
\caption{Defect structure of system A-2D: The pictures show results at timestep $t=5.12\e{6}$ (top) and $1.024\e{7}$ (bottom) with an imposed shear rate $\dot{\gamma}=1.25\cdot10^{-5}$. The system has been pre-sheared at various higher, subsequently lowered shear rates ranging from $\dot{\gamma}=2\e{-4}$ to the above value (see Tab.\ref{tab1}). Sites where the alignment function $X$ was smaller than $0.8$ are highlighted red.}
\label{fig3}
\end{figure}

Defect annihilation to this large extent is only found below a specific shear rate.
A prerequisite seems to be that the defects can smootly approach each other and cancel, which is prevented by the buckling instability above a criticl shear rate.
Then the system shows a tendency to trip over itself, which results in the creation of new defects.
The critical value differs in systems A-2D and B-2D and is related to the magnitude of the order parameter and the distance from the critical point.
It is roughly between $\dot{\gamma}=2.5-5\e{-5}$ in A-2D and around $\dot{\gamma}=1-2\e{-4}$ in B-2D.\\

\begin{figure}[htp]
\centering
\hspace*{-0.1cm}\includegraphics[angle=0,width=0.5\textwidth]{phi_run704_400.jpg}
\includegraphics[angle=0,width=0.49\textwidth]{shear_str_run704_400.jpg}
\includegraphics[angle=0,width=0.49\textwidth]{N1_run704_400.jpg}
\caption{Order parameter $\phi({\mathbf r})$ (top), shear stress $\sigma_{xy}({\mathbf r})$ (centre) and first normal stress differences $N_1({\mathbf r})$ (bottom) of system A-2D at $t=4\e{5}$}
\label{fig4}
\end{figure}

In Fig. \ref{fig4} we suggest a relation between the inhomogeneous morphology in the flowing state and typical rheological quantities.
The pictures show the order parameter $\phi({\bm r})$, shear stress density $\sigma_{xy}({\bm r})$ and first normal stress difference density $N1({\bm r})=\sigma_{yy}({\bm r})-\sigma_{xx}({\bm r})$ of system A-2D in steady shear flow at shear rate $\dot{\gamma}=1\e{-4}$.
In regions where the lamellar width is larger than in the equilibrium state shear stresses and first normal stress differences show low positive or negative values.
Large positive values of both quantities coincide with regions where the lamellae are under compression.
Peak values in LBU are in the region of $-2\e{-4}$ and $3\e{-4}$ for shear stress and  $-3.5\e{-4}$ and $7\e{-4}$ for the normal stress differences.
Despite being locall negative, the ensemble averages of these quantities are always positive.
Very similar results with smaller peak magnitudes are obtained for system B-2D.

\begin{figure}[htp]
\centering
\includegraphics[angle=0,width=0.33\textwidth]{ck_run703_320.jpg}
\includegraphics[angle=0,width=0.33\textwidth]{ck_run704_500.jpg}
\includegraphics[angle=0,width=0.33\textwidth]{ck_run705_1280.jpg}
\caption{Structure factor $C({\bm k})$ of system A-2D: Shown are states after equilibration at $t=3.2\e{5}$ (left), at $\dot{\gamma}=1\cdot10^{-4}$ and $t=5\e{5}$ (centre) and after switch-off at $t=1.28\e{6}$ (right). The sections shown correspond to wave vectors on the interval $k_x, k_y \in [-\pi/2 l,\pi/2 l]$.}
\label{fig5}
\end{figure}

Fig. \ref{fig5} shows structure factors $C({\bm k})$ for the same states as Fig. \ref{fig1}.
Initially isotropic after spinodal decomposition and equilibration, $C({\bm k})$ has a distinct peak at $|{\bm k}|\simeq \pm0.2\pi/l$.
In steady shear flow the structure factor develops two distinct but rather broad peaks at $k_x\simeq\mp 0.012\pi/l, k_y\simeq\pm 0.22 \pi/l$.
It is interesting to see that the inhomogenous morphology of the lamellar structure is also clearly encoded in the structure factor. 
The Fourier-vector at the peak has a small component in x-direction, indicating the average tilt of the lamellae away from the flow direction in the direction of extensional flow.
The position of the peak is almost constant throughout the entire simulation and the average lamellar spacing decreases from about 10 lattice sites in equilibrium to about 8.5 lattice sites.
If the shear flow is switched off the system adopts again its equilibrium value and the small tilt of the signal vanishes completely, indicating that this is a purely dynamical phenomenon. 
The intensity of the peak is slightly larger due to the fact that the interfaces remain oriented along the flow direction.\\

Fig. \ref{fig6} shows the  complete time evolution of the defects density $\rho_D$ of system A-2D for a sequence of decreasing shear rates starting at $\gd=2\e{-4}$ and ranging down to $\gd=2.5\e{-5}$.
The runs with the two highest shear rates were started from the equilibrium configuration, whereas all others were pre-sheared using the final configuration of the preceding run as initial configuration in order to save computational rexources.
The total strain between the beginning and end of every individual run was at least $10^{5}\%$.
Interestingly, while stepping down in shear rate both systems A-2D and B-2D anneal defects almost instantaneously.
If the shear rate is turned up again (not shown), the lamellae buckle and new defects recur immediately.

\begin{figure}[htp]
\centering
\includegraphics[angle=0,width=0.5\textwidth]{defect_density_5e-4.pdf}
\caption{Defect density $\rho_D$ vs time: The main picture shows data for system A-2D of a sequence of simulations with consecutively lowered shear rate $\dot{\gamma}=2\e{-4}$ (red solid), $1\e{-4}$ (green long-dashed),  $5\e{-5}$ (blue short-dashed),  $2.5\e{-5}$ (magenta dotted) and  $1.25\e{-5}$ (cyan dashed-dotted). Horizontal lines indicate time averages of the particular quantities, where we assume the system has reached a steady state. The inset gives data for the system A-3D and shear rates $\dot{\gamma}=2\cdot10^{-4}, 1\cdot10^{-4}$ and $5\cdot10^{-5}$.} 
\label{fig6}
\end{figure}

@@

This feature which appears somewhat erratic at first has actually a simple explanation. 
The time sequence of the defect density in Fig. \ref{fig6} reveals that the rise in $N_1$ is correlated with a drop in $\rho_D$ and linked to the defect annihilation.
When two defects meet and cancel the lamellar width increases locally by a certain amount.
This causes frustration in the system as the incommensurability between the layers is not large enough to create a new lamellar layer and allow the system to snap back towards its equilibrium width.
Hence the normal stresses $\sigma_{xx}$ become increasingly smaller and may even attain negative values, leading to gradually larger values of $N_1$.
This explanation is further sustained by the fact that only $\sigma_{xx}$ seems to evolve anomalously slowly and all other components of the stress tensor reach their steady state values more quickly.
We expect that these anomalous trends become considerably smaller in larger systems.
System B-2D behaves differently in the sense that $N_1$ does not follow this trend but rather continues to fall for lower shear rates. 
This is likely to be a direct consequence of the lower order parameter in system B-2D and its softness with respect to system A-2D, which gives rise to positive normal stresses $\sigma_{xx}$ preventing $N_1$ from increasing.\\
In Fig. \ref{fig11} the shear rate dependence of $\sigma_{xy}$ and $N_1$ is shown. 
System A-2D gives rise to shear thinning with the average shear stress of systems A-2D following a power law $\sigma_{xy}=\dot{\gamma}^m$ with exponent $m\simeq 0.87$.
For system B-2D the best fit is actually obtained for an exponent $m=1$, indicating more or less constant viscosity and no pronounced shear thinning.
This is because due to the decreasing number of defects at shear rates below $\dot{\gamma}\le2\e{-4}$ the chemical stress $P^{chem}_{xy}$ almost vanishes and the total stress is dominated by the hydrodynamic part.\\
